Arrows for Parallel Computations

### 2017-11-28 paper/content/abstract.tex
tmp

Additionally, using these parallel Arrows (PArrows) has the benefit of being portable across multiple parallel Haskell implementations.
Furthermore, as each parallel computation is an Arrow, PArrows can be readily composed and transformed as such.
In order to allow for more sophisticated communication schemes between computation nodes in distributed systems we introduce the concept of Futures to wrap existing direct communication concepts in backends.

To show that PArrows have similar expressiveness to existing parallel languages, we also implement several parallel skeletons. 
Benchmarks show that our framework does not induce any notable overhead
performance-wise. We finally conclude that Arrows turn out to be a useful tool for composing parallel programs and that our particular approach results in programs that are portable accross multiple backends.
\olcomment{Summarize conclusions}

>>>>

Additionally, using these parallel Arrows (PArrows) has the benefit of being portable across multiple parallel Haskell implementations.
Furthermore, as each parallel computation is an Arrow, PArrows can be readily composed and transformed as such.
In order to allow for more sophisticated communication schemes between computation nodes in distributed systems we utilise the concept of Futures to wrap existing direct communication concepts in backends.

To show that PArrows have similar expressive power as existing parallel languages, we also implement several parallel skeletons. 
Benchmarks show that our framework does not induce any notable overhead
performance-wise. We finally conclude that Arrows turn out to be a useful tool for composing parallel programs and that our particular approach results in programs that are portable accross multiple backends.
\olcomment{Summarize conclusions}

### 2017-11-28 paper/content/benchmarks.tex
minor
tmp

Apart from Eden, all benchmarks and libaries were compiled with Stack's \footnote{see \url{https://www.haskellstack.org/}} lts-7.1 GHC compiler which is equivalent to a standard GHC 8.0.1 with the base package in version 4.9.0.0. Stack itself was used in version 1.3.2. For GpH in its Multicore variant we used the the parallel cabal package in version 3.2.1.0, while for the |Par| monad we used monad-par in version 0.3.4.8. For all Eden tests, we used its GHC-Eden compiler in version 7.8.2 together with OpenMPI 1.6.5.

Furtermore, all benchmarks were done with help of the bench \footnote{see \url{https://hackage.haskell.org/package/bench}} tool in version 1.0.2 which uses criterion (>=1.1.1.0 \&\& < 1.2) internally. All runtime data (mean runtime, max stddev, etc.) was collected with this tool if not mentioned otherwise.

@@

\begin{table}
\centering
%% something was wrong with separators in table
\renewcommand{\tabcolsep}{0.5em}
\begin{tabular}{lccll}
\toprule
Name & Area & Type & Origin & Source \\
\midrule
\rmtest & Mathematics & \ensuremath{\Varid{parMap}\mathbin{+}\Varid{reduce}} & Eden & \citet{Lobachev2012}\\
\jacobitest & Mathematics & \ensuremath{\Varid{workpool}\mathbin{+}\Varid{reduce}} & Eden & \citet{Lobachev2012}\\
\torustest & Mathematics & \ensuremath{\Varid{torus}} & Eden & \citet{Eden:SkeletonBookChapter02}\\
\sudokutest & Puzzle & \ensuremath{\Varid{parMap}} & \ensuremath{\Conid{Par}} Monad & \citet{par-monad} 
\tablefootnote{actual code from: http://community.haskell.org/\~simonmar/par-tutorial.pdf and https://github.com/simonmar/parconc-examples}\\
\bottomrule
\end{tabular}
\label{tab:benches}
\caption{The benchmarks we use in this paper.}
\end{table}

>>>>

Apart from Eden, all benchmarks and libaries were compiled with Stack's\footnote{see \url{https://www.haskellstack.org/}} lts-7.1 GHC compiler which is equivalent to a standard GHC 8.0.1 with the base package in version 4.9.0.0. Stack itself was used in version 1.3.2. For GpH in its Multicore variant we used the the parallel package in version 3.2.1.0\footnote{see \url{https://hackage.haskell.org/package/parallel-3.2.1.0}}, while for the |Par| monad we used monad-par in version 0.3.4.8\footnote{see \url{https://hackage.haskell.org/package/monad-par-0.3.4.8}}. For all Eden tests, we used its GHC-Eden compiler in version 7.8.2\footnote{see \url{http://www.mathematik.uni-marburg.de/~eden/?content=build_eden_7_&navi=build}} together with OpenMPI 1.6.5\footnote{see \url{https://www.open-mpi.org/software/ompi/v1.6/}}.

Furtermore, all benchmarks were done with help of the bench\footnote{see \url{https://hackage.haskell.org/package/bench}} tool in version 1.0.2 which uses criterion (>=1.1.1.0 \&\& < 1.2)\footnote{see \url{https://hackage.haskell.org/package/criterion-1.1.1.0}} internally. All runtime data (mean runtime, max stddev, etc.) was collected with this tool if not mentioned otherwise.

@@

\begin{table}
\centering
\caption{The benchmarks we use in this paper.}
%% something was wrong with separators in table
\renewcommand{\tabcolsep}{0.5em}
\begin{tabular}{lccll}
\toprule
Name & Area & Type & Origin & Source \\
\midrule
\rmtest & Mathematics & \ensuremath{\Varid{parMap}\mathbin{+}\Varid{reduce}} & Eden & \citet{Lobachev2012}\\
\jacobitest & Mathematics & \ensuremath{\Varid{workpool}\mathbin{+}\Varid{reduce}} & Eden & \citet{Lobachev2012}\\
\torustest & Mathematics & \ensuremath{\Varid{torus}} & Eden & \citet{Eden:SkeletonBookChapter02}\\
\sudokutest & Puzzle & \ensuremath{\Varid{parMap}} & \ensuremath{\Conid{Par}} Monad & \citet{par-monad}\tablefootnote{actual code from: \url{http://community.haskell.org/\~simonmar/par-tutorial.pdf and https://github.com/simonmar/parconc-examples}}\\
\bottomrule
\end{tabular}
\label{tab:benches}
\end{table}

### 2017-11-28 paper/content/bestAndWorstBenchmarks.tex
tmp

\newcommand{\meanOverheadTEST}{
\begin{table}[]
\centering
\caption{Overhead in the benchmarks}
\label{tab:meanOverhead}
\captionsetup[subtable]{justification=centering}
\subfloat[\sudokutest 1000]{
	\begin{minipage}[t]{\linewidth}%
    \centering%
    \begingroup\catcode`"=9
	\begin{tabular}{C{2cm} C{3cm} C{1.7cm} C{2.3cm}}
		\thead{vs.}             & \thead{Benchmark} & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \toprule
		\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-1000.csv}{}{\vs & \benchmark \params & \meanOverhead & \maxStdDevForOverhead \\}
		\midrule
	\end{tabular}
	\endgroup
	\label{sudokuMeanFactor-1000}
	\end{minipage}
}
\caption{Overhead in the benchmarks}
\end{table}
}

\newcommand{\meanOverhead}{

@@

@@

@@

\meanOverhead

>>>>

@@

\sisetup{
	table-number-alignment = center

@@

	\begin{tabular}{S S[table-text-alignment = center] S[table-text-alignment = center]}
		vs.             & mean overhead & maximum normalized stdDev \\ \hline

@@

### 2017-11-28 paper/content/mapSkeletons.tex
reworking contributions
tmp

%%% \FloatBarrier
\section{Map-based Skeletons}
\label{sec:map-skeletons}
Now we have developed Parallel Arrows far enough to define some useful algorithmic skeletons that abstract typical parallel computations.

>>>>

\section{Skeletons}
\label{sec:skeletons}
Now we have developed Parallel Arrows far enough to define some useful algorithmic skeletons that abstract typical parallel computations. While there are many possible skeletons to implement, we regard in detail here only some |map|-based and topological skeletons to demonstrate the power of PArrows.
%%% \FloatBarrier
\subsection{|map|-based Skeletons}
\label{sec:map-skeletons}
We start with |map|-based skeletons. The essential differences between the skeletons presented here are in terms of order of evaluation and work distribution but still provide the same output of a standard |map|.

### 2017-11-28 paper/content/motivation.tex
tmp

%Our Arrows-based interface is a high-level one.
We provide an Arrows-based type class and implementations for the three above mentioned parallel Haskells.
Instead of 
introducing a new low-level parallel backend in order to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface with varying implementations in
the existing parallel Haskells. Thus, we not only define a parallel programming interface in a
novel manner - we tame the zoo of parallel Haskells. We provide a
common \ptcomment{quantify, e.g. less than 10 \%} \mbcomment{we do this in the contributions section of the Introduction, enough?}, very low-penalty programming interface that allows to switch
the parallel backends at will. The induced penalty was max. 8.3 \% in our measurements. Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.

\paragraph{Contributions}
%
%\olcomment{HIT HERE REALLY STRONG}
%
%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
We propose a new Arrow-based encoding for parallelism based on a new Arrow combinator |parEvalN :: [arr a b] -> arr [a] [b]| in Section \ref{sec:parallel-arrows}, which converts a list of Arrows into a new parallel Arrow. A parallel Arrow is still an Arrow, hence the resulting parallel Arrow can still be used in the same way as a potential sequential version.

We evaluate the expressive power of Arrow formalism in the context of parallel programming.

\begin{itemize}
\item We introduce a parallel evaluation formalism using Arrows. One big advantage is that we do not introduce any new types (Sec.~\ref{sec:parallel-arrows}) in contrast Monad solutions such as |Par| Monad. This behaviour encourages better composability.
\item We utilitze multiple backends -- currently a GpH, a |Par| Monad, and Eden. We do not reimplement all the parallel internals, as we host this functionality in the |ArrowParallel| type class, which abstracts all parallel implementation logic. The backends can easily be swapped, so we are not bound to any specific one.

So as an example, during development, we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant. (Sec.~\ref{sec:parallel-arrows})
\item We extend our PArrows formalism with |Future|s. Our goal here is to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data (|RD|). Direct communication is useful in a distributed memory setting because they allow for inter-node communication without blocking the master-node. (Sec.~\ref{sec:futures})
\item It is possible to define algorithmic skeletons with PArrows (Sec.~\ref{sec:skeletons})
\item Finally, we demonstrate that Arrow parallelism is a viable alternative to existing approaches. It introduces only low overhead (Sec.~\ref{sec:benchmarks}).
\end{itemize}

>>>>

%Our Arrows-based interface is a high-level one.
We provide an Arrows-based type class and implementations for the three above mentioned parallel Haskells.
Instead of 
introducing a new low-level parallel backend in order to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface with varying implementations in
the existing parallel Haskells. Thus, we not only define a parallel programming interface in a
novel manner -- we tame the zoo of parallel Haskells. We provide a
common, very low-penalty programming interface that allows to switch
the parallel backends at will. The induced penalty was max. 8.3 \% in our measurements \mbcomment{, and typically under X \%}. Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.

\paragraph{Contributions}
%
%\olcomment{HIT HERE REALLY STRONG}
%
%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
We propose an Arrow-based encoding for parallelism based on a new Arrow combinator |parEvalN :: [arr a b] -> arr [a] [b]| in Section \ref{sec:parallel-arrows}, which converts a list of Arrows into a new parallel Arrow. A parallel Arrow is still an Arrow, hence the resulting parallel Arrow can still be used in the same way as a potential sequential version.

In this paper evaluate the expressive power of our Arrow formalism in the context of parallel programming.

\begin{itemize}
\item We introduce a parallel evaluation formalism using Arrows. One big advantage is that we do not have to introduce any new types (Sec.~\ref{sec:parallel-arrows}) in contrast to Monad solutions such as the |Par| Monad. This behaviour encourages better composability.
\item We utilitze multiple backends -- currently a GpH, a |Par| Monad, and Eden. We do not reimplement all the parallel internals, as we host this functionality in the |ArrowParallel| type class, which abstracts all parallel implementation logic. The backends can easily be swapped, so we are not bound to any specific one.

So as an example, during development, we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant. (Sec.~\ref{sec:parallel-arrows})
\item We extend our PArrows formalism with |Future|s. Our goal here is to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data (|RD|). Direct communication is useful in a distributed memory setting because it allows for inter-node communication without blocking the master-node. (Sec.~\ref{sec:futures})
\item It is possible to define algorithmic skeletons with PArrows (Sec.~\ref{sec:skeletons}). All our benchmarks were skeleton-based.
\item Finally, we practically demonstrate that Arrow parallelism is a viable alternative to existing approaches. It introduces only low performance overhead (Sec.~\ref{sec:benchmarks}).
\end{itemize}
